---
title: "MLB Statcast: A Hierarchical Bayesian Stuff+ Model"
author: "Trey Chase"
format:
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 4.5
editor: visual
---

# Overview

This notebook builds a **Stuff+** metric for individual pitches from Statcast
data. Everything is in R: the scrape and reduction with `baseballr` and the
tidyverse, the Bayesian models with `brms` on top of Stan.

1.  Scrape pitch-level Statcast data and reduce it to a pitcher × pitch-type
    table of average release/movement mechanics, whiff rate, and **cumulative
    run value**.
2.  Fit a **hierarchical Bayesian beta-binomial model** of whiff propensity on
    mechanics, with a varying intercept *and* varying slopes by `pitch_name`,
    so each pitch type gets its own posterior.
3.  Define **Stuff+** as the standardized z-score of a pitcher's posterior
    latent whiff propensity relative to the posterior for *his own pitch type*,
    on the conventional 100 ± 10 scale.
4.  **Test the linearity assumption** of the logit-linear predictor. Where it
    fails, do targeted EDA and fit a second MCMC model with penalized splines.
5.  Run full MCMC diagnostics, then **select the model by out-of-sample
    $R^2$ and MSE against cumulative run value**.

```{r}
#| label: setup
#| message: false
#| warning: false

library(baseballr)
library(tidyverse)
library(brms)
library(posterior)
library(bayesplot)
library(loo)
library(splines)

knitr::opts_chunk$set(message = FALSE, warning = FALSE)

options(mc.cores = min(4, parallel::detectCores()))

# brms evaluates the beta-binomial log-likelihood through extraDistr. It is a
# suggested rather than a hard dependency, so it is easy to have a working brms
# that samples fine and only fails later at loo(). Worse, when loo() runs in
# parallel the missing-package error surfaces from the worker as the unrelated
# "Error in exp(x): non-numeric argument to mathematical function", which sends
# you looking in the wrong place. Fail early with the real message instead.
if (!requireNamespace("extraDistr", quietly = TRUE)) {
  stop("beta_binomial() log-likelihood needs the 'extraDistr' package: ",
       "install.packages('extraDistr')")
}

# Compiled Stan binaries are cached, so re-running does not recompile.
rstan::rstan_options(auto_write = TRUE)

RANDOM_SEED <- 42
set.seed(RANDOM_SEED)

# One place to change the sampler settings. adapt_delta is raised from the
# default 0.8 because hierarchical models with varying slopes routinely produce
# divergent transitions at the default.
CHAINS      <- 4
ITER        <- 4000   # includes warmup
WARMUP      <- 2000
ADAPT_DELTA <- 0.95

DATA_DIR <- "data"
dir.create(DATA_DIR, showWarnings = FALSE, recursive = TRUE)
```

# Data acquisition

```{r}
#| label: scrape
#| cache: true

date_starts <- seq(as.Date("2026-04-10"), Sys.Date(), by = "7 days")
date_ends   <- pmin(date_starts + 6, Sys.Date())

data <- map2_dfr(
  date_starts, date_ends,
  ~ scrape_statcast_savant(
      start_date  = .x,
      end_date    = .y,
      player_type = "pitcher"
    )
)

nrow(data)
```

# Pitch-level cleaning

Three changes relative to a naive `select() |> distinct()`:

**Run value is carried through.** `delta_run_exp` is Statcast's change in run
expectancy attributable to each pitch (negative = good for the pitcher). It is
the outcome the model is ultimately selected against, so it has to survive the
reduction.

**De-duplication uses the true pitch key.** `distinct()` over the *value*
columns silently collapses two genuinely different pitches that happen to share
rounded measurements. `game_pk` + `at_bat_number` + `pitch_number` uniquely
identifies a pitch, so dedupe on that instead.

**Horizontal quantities are mirrored for left-handers.** `pfx_x` and
`release_pos_x` are recorded in the catcher's frame, so a lefty's arm-side run
carries the opposite sign to a righty's. Averaging the two together pushes group
means toward zero and makes the coefficient on `avg_pfx_x` close to meaningless.
Flipping the sign for LHP puts every pitch in a **pitcher-centric** frame:
positive `pfx_x` is arm-side, negative is glove-side.

```{r}
#| label: pitch-df

# `arm_angle` was added to the Statcast feed relatively recently; older
# baseballr versions do not return it and `select()` would abort. Create it as
# missing instead and let the missingness report and VIF screen decide what to
# do with it. `delta_run_exp` is not optional -- it is the model-selection
# target -- so fail loudly if it is absent.
if (!"arm_angle" %in% names(data)) {
  warning("arm_angle not returned by this baseballr version; filling with NA")
  data$arm_angle <- NA_real_
}
stopifnot("delta_run_exp missing - update baseballr" =
            "delta_run_exp" %in% names(data))

pitch_df <- data |>
  select(
    game_pk, at_bat_number, pitch_number,
    player_name, pitcher, p_throws, stand,
    pitch_name,
    release_speed, release_pos_x, release_pos_z,
    release_spin_rate, release_extension, spin_axis, arm_angle,
    pfx_x, pfx_z, plate_x, plate_z,
    description, delta_run_exp
  ) |>
  distinct(game_pk, at_bat_number, pitch_number, .keep_all = TRUE)

# Drop pitch types that are too rare or too heterogeneous to model.
pitch_df <- pitch_df |>
  filter(!pitch_name %in% c(
    "Forkball", "Pitch Out", "Eephus", "Knuckleball",
    "Other", "Slow Curve", "Unknown", "NA"
  ), !is.na(pitch_name), pitch_name != "")

# Pitcher-centric horizontal frame.
pitch_df <- pitch_df |>
  mutate(
    pfx_x         = if_else(p_throws == "L", -pfx_x, pfx_x),
    release_pos_x = if_else(p_throws == "L", -release_pos_x, release_pos_x),
    spin_axis     = if_else(p_throws == "L", (360 - spin_axis) %% 360, spin_axis)
  )

nrow(pitch_df)
```

# Exploratory data analysis

```{r}
#| label: eda-velo
ggplot(pitch_df, aes(x = release_speed, fill = pitch_name)) +
  geom_histogram(bins = 30, color = "white") +
  facet_wrap(~ pitch_name, scales = "free") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Release Speed (mph)", y = "Count",
       title = "Release Speed Distribution by Pitch Type")
```

```{r}
#| label: eda-release
ggplot(pitch_df, aes(x = release_pos_x, y = release_pos_z, color = p_throws)) +
  geom_point(alpha = 0.25, size = 0.5) +
  labs(
    x = "Release Position X (ft, arm-side positive)",
    y = "Release Position Z (ft)",
    title = "Release Position by Pitcher Handedness (mirrored for LHP)",
    color = "Throws"
  ) +
  theme_minimal()
```

After mirroring, the two handedness clouds overlap instead of sitting on
opposite sides of the plate — which is the point: a left-hander's slider and a
right-hander's slider now occupy the same region of mechanics space and can
share a pitch-type-level prior.

```{r}
#| label: eda-movement
ggplot(pitch_df, aes(x = pfx_x * 12, y = pfx_z * 12, color = pitch_name)) +
  geom_point(alpha = 0.15, size = 0.4) +
  coord_fixed() +
  labs(x = "Horizontal break (in, arm-side positive)",
       y = "Induced vertical break (in)",
       title = "Movement Profile by Pitch Type", color = NULL) +
  theme_minimal()
```

# Aggregation to pitcher × pitch type

```{r}
#| label: summarise

swing_descriptions <- c(
  "foul", "foul_tip", "hit_into_play", "swinging_strike",
  "swinging_strike_blocked", "missed_bunt", "foul_bunt"
)
whiff_descriptions <- c("swinging_strike", "swinging_strike_blocked", "missed_bunt")

pitch_df <- pitch_df |>
  mutate(
    is_swing = description %in% swing_descriptions,
    is_whiff = description %in% whiff_descriptions
  )

pitcher_summary <- pitch_df |>
  group_by(pitch_name, player_name) |>
  summarise(
    avg_release_speed     = mean(release_speed, na.rm = TRUE),
    avg_release_pos_x     = mean(release_pos_x, na.rm = TRUE),
    avg_release_pos_z     = mean(release_pos_z, na.rm = TRUE),
    avg_release_spin_rate = mean(release_spin_rate, na.rm = TRUE),
    avg_release_extension = mean(release_extension, na.rm = TRUE),
    avg_arm_angle         = mean(arm_angle, na.rm = TRUE),
    avg_pfx_x             = mean(pfx_x, na.rm = TRUE),
    avg_pfx_z             = mean(pfx_z, na.rm = TRUE),
    n_swings  = sum(is_swing, na.rm = TRUE),
    n_whiffs  = sum(is_whiff, na.rm = TRUE),
    n_pitches = n(),
    run_value = sum(delta_run_exp, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_pitches >= 20, n_swings >= 10) |>
  mutate(
    whiff_rate = n_whiffs / n_swings,
    rv_per_100 = 100 * run_value / n_pitches
  ) |>
  arrange(pitch_name, player_name)

MECH_VARS <- c(
  "avg_release_speed", "avg_release_pos_x", "avg_release_pos_z",
  "avg_release_spin_rate", "avg_release_extension", "avg_arm_angle",
  "avg_pfx_x", "avg_pfx_z"
)

write_csv(pitcher_summary, file.path(DATA_DIR, "pitcher_summary.csv"))
nrow(pitcher_summary)
```

```{r}
#| label: whiff-hist
wr_mean <- mean(pitcher_summary$whiff_rate, na.rm = TRUE)
wr_var  <- var(pitcher_summary$whiff_rate, na.rm = TRUE)

ggplot(pitcher_summary, aes(x = whiff_rate)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = wr_mean, color = "red", linetype = "dashed") +
  annotate("text", x = wr_mean, y = Inf,
           label = paste0("Mean = ", round(wr_mean, 3),
                          "\nVar = ", round(wr_var, 3)),
           vjust = 1.5, hjust = -0.1, color = "red") +
  labs(x = "Whiff Rate (Whiffs / Swings)", y = "Count",
       title = "Whiff Rate by Pitcher and Pitch Type (20+ pitches, 10+ swings)") +
  theme_minimal()
```

The binomial variance implied by the mean whiff rate is far smaller than the
observed variance across pitcher-pitch rows, so the counts are **overdispersed**
relative to a binomial. That is the substantive reason for `beta_binomial()`
below rather than `binomial()`.

```{r}
#| label: rv-hist
ggplot(pitcher_summary, aes(x = run_value)) +
  geom_histogram(bins = 40, fill = "darkorange", color = "white") +
  labs(x = "Cumulative run value (runs, negative favours pitcher)",
       y = "Count", title = "Cumulative Run Value by Pitcher and Pitch Type") +
  theme_minimal()
```

# Data hygiene before modelling

`mean(x, na.rm = TRUE)` returns `NaN` when a group is entirely missing, so a
pitcher can clear the `n_pitches >= 20` filter and still carry `NaN` mechanics.
`arm_angle` is the usual culprit — it is a relatively new Statcast field and is
absent for some pitchers entirely. `brm()` would drop those rows silently; they
are dropped here explicitly instead, **before** standardizing, so the means and
standard deviations used to z-score the predictors are computed on exactly the
rows that get modelled.

```{r}
#| label: missingness

# NaN and +/-Inf both count as missing here.
pitcher_summary <- pitcher_summary |>
  mutate(across(all_of(MECH_VARS), ~ if_else(is.finite(.x), .x, NA_real_)))

missing_report <- pitcher_summary |>
  summarise(across(all_of(MECH_VARS), ~ sum(is.na(.x)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  mutate(pct = round(100 * n_missing / nrow(pitcher_summary), 2)) |>
  arrange(desc(n_missing))
print(missing_report)

incomplete <- pitcher_summary |>
  filter(if_any(all_of(MECH_VARS), is.na))
cat(sprintf("\n%d of %d rows incomplete (%.2f%%) -> dropped\n",
            nrow(incomplete), nrow(pitcher_summary),
            100 * nrow(incomplete) / nrow(pitcher_summary)))
print(count(incomplete, pitch_name, sort = TRUE))

model_df <- pitcher_summary |> drop_na(all_of(MECH_VARS))
```

Because the dropped rows are concentrated in whichever pitch types happen to
lack `arm_angle` coverage rather than being spread at random, this is
missing-not-completely-at-random. At well under 2% of rows it is not worth
imputing, but it *is* worth checking that no pitch type loses a meaningful share
of its pitchers — the group-level terms below need enough pitchers per type to
estimate a variance.

```{r}
#| label: group-sizes
MIN_PITCHERS_PER_TYPE <- 25

sizes <- full_join(
  count(pitcher_summary, pitch_name, name = "before"),
  count(model_df, pitch_name, name = "after"),
  by = "pitch_name"
) |>
  mutate(across(c(before, after), ~ replace_na(.x, 0L)),
         pct_lost = round(100 * (1 - after / before), 1)) |>
  arrange(desc(after))
print(sizes)

# A pitch type with only a handful of pitchers cannot support its own slopes; it
# would be shrunk almost entirely to the population mean anyway, and it
# destabilises the group-level variances.
keep_types <- sizes$pitch_name[sizes$after >= MIN_PITCHERS_PER_TYPE]
dropped_types <- setdiff(sizes$pitch_name, keep_types)
if (length(dropped_types)) {
  cat("dropping sparse pitch types:", paste(dropped_types, collapse = ", "), "\n")
}
model_df <- model_df |>
  filter(pitch_name %in% keep_types) |>
  mutate(pitch_name = factor(pitch_name))

cat(sprintf("modelling rows: %d, pitch types: %d\n",
            nrow(model_df), nlevels(model_df$pitch_name)))
```

## Collinearity screen

Before fitting, check whether the eight mechanics are separately identifiable.
`arm_angle` is a particular concern: Statcast derives it from the release point
and the pitcher's shoulder position, so it is close to a deterministic function
of `release_pos_x` and `release_pos_z`. Slopes on mutually collinear predictors
are individually unstable even when the model's predictions are fine — and
unstable slopes are exactly what Stuff+ is built out of.

The model already carries a varying intercept for `pitch_name`, so between-type
differences are absorbed there and do not destabilise the slopes. What matters
is collinearity **within** a pitch type. Pooled VIF is dominated by the fact that
sliders are simply slower and spin differently than four-seamers, and would
throw away genuinely informative movement variables for the wrong reason.

```{r}
#| label: vif

# VIF_j = 1 / (1 - R^2_j) from regressing predictor j on the others. Written out
# rather than pulled from car::vif so the notebook needs no extra dependency.
vif_table <- function(df, cols) {
  vapply(cols, function(j) {
    others <- setdiff(cols, j)
    r2 <- summary(lm(reformulate(others, response = j), data = df))$r.squared
    1 / (1 - r2)
  }, numeric(1)) |>
    sort(decreasing = TRUE) |>
    round(2)
}

within_group_center <- function(df, cols, group = "pitch_name") {
  df |>
    group_by(.data[[group]]) |>
    mutate(across(all_of(cols), ~ .x - mean(.x))) |>
    ungroup() |>
    select(all_of(cols))
}

centered <- within_group_center(model_df, MECH_VARS)

cat("pooled correlation (between + within pitch type)\n")
print(round(cor(model_df[MECH_VARS]), 2))
cat("\nwithin-pitch-type correlation\n")
print(round(cor(centered), 2))
cat("\npooled VIF (misleading here)\n")
print(vif_table(model_df, MECH_VARS))
cat("\nwithin-pitch-type VIF (the one that matters)\n")
print(vif_table(centered, MECH_VARS))
```

```{r}
#| label: vif-prune
VIF_THRESHOLD <- 5

prune_by_vif <- function(df, cols, threshold = VIF_THRESHOLD) {
  dropped <- character(0)
  while (length(cols) > 1) {
    v <- vif_table(df, cols)
    if (v[1] < threshold) break
    cat(sprintf("dropped %s (within-type VIF = %.2f)\n", names(v)[1], v[1]))
    dropped <- c(dropped, names(v)[1])
    cols <- setdiff(cols, names(v)[1])
  }
  cols
}

PRED_VARS <- prune_by_vif(centered, MECH_VARS)
cat("\nretained predictors:", paste(PRED_VARS, collapse = ", "), "\n")
print(vif_table(centered, PRED_VARS))
```

## Standardization

```{r}
#| label: standardize
std_stats <- model_df |>
  summarise(across(all_of(PRED_VARS), list(mean = mean, sd = sd))) |>
  pivot_longer(everything(),
               names_to = c("var", ".value"),
               names_pattern = "^(.*)_(mean|sd)$")
print(std_stats |> mutate(across(c(mean, sd), ~ round(.x, 3))))

model_df <- model_df |>
  mutate(across(all_of(PRED_VARS), ~ as.numeric(scale(.x))))

stopifnot(all(model_df$n_whiffs <= model_df$n_swings))
stopifnot(!anyNA(model_df[PRED_VARS]))
```

# Model 1 — hierarchical linear beta-binomial

Swings are the trials and whiffs the successes. A plain binomial would assume
the only variation in whiff rate at fixed mechanics is sampling noise; the
histogram above shows far more spread than that, so `beta_binomial()` carries an
extra dispersion parameter `phi`.

The varying intercept **and** varying slopes by `pitch_name` are what make this
hierarchical: velocity should matter differently for a four-seamer than for a
changeup, and each pitch type gets a partially-pooled posterior of its own
rather than being forced onto one league-wide slope.

Note the `||` rather than `|`. brms defaults to *correlated* group-level terms,
which here would mean estimating a full correlation matrix over the intercept
and every slope — dozens of correlation parameters from a handful of pitch
types. That is not identifiable from this much data and shows up as divergent
transitions and $\hat{R}$ warnings. `||` drops the correlations and keeps one
variance per term, which is what the data can actually support.

```{r}
#| label: m1-formula
hierarchical_formula <- function(terms, group = "pitch_name") {
  fixed <- paste(terms, collapse = " + ")
  as.formula(sprintf(
    "n_whiffs | trials(n_swings) ~ %s + (1 + %s || %s)", fixed, fixed, group
  ))
}

f1 <- hierarchical_formula(PRED_VARS)
print(f1)

# Weakly informative priors on the logit scale. Predictors are standardized, so
# a slope of 1 means a one-SD change in mechanics moves the log-odds of a whiff
# by 1 -- already large. Normal(0, 1) is weakly informative here, not restrictive.
priors1 <- c(
  prior(normal(0, 1.5), class = "Intercept"),
  prior(normal(0, 1),   class = "b"),
  prior(exponential(1), class = "sd"),
  prior(gamma(0.01, 0.01), class = "phi")
)
```

```{r}
#| label: m1-fit
#| cache: true
m1 <- brm(
  f1, data = model_df, family = beta_binomial(),
  prior = priors1,
  chains = CHAINS, iter = ITER, warmup = WARMUP,
  control = list(adapt_delta = ADAPT_DELTA),
  seed = RANDOM_SEED, refresh = 0
)
summary(m1)
```

# MCMC diagnostics

A posterior is only worth interpreting if the sampler actually explored it. The
checks below are the standard battery: split $\hat{R}$ (should be < 1.01), bulk
and tail effective sample size (should comfortably exceed 400 per parameter),
divergent transitions (should be zero), tree-depth saturation, and E-BFMI (below
~0.3 indicates trouble with the energy distribution).

```{r}
#| label: diagnostics
mcmc_report <- function(fit, name) {
  s <- summarise_draws(as_draws_array(fit), "rhat", "ess_bulk", "ess_tail") |>
    filter(variable != "lp__")
  np <- nuts_params(fit)
  div <- sum(np$Value[np$Parameter == "divergent__"])
  td  <- np$Value[np$Parameter == "treedepth__"]
  n_draws <- (ITER - WARMUP) * CHAINS
  bfmi <- tryCatch(rstan::get_bfmi(fit$fit), error = function(e) NA_real_)
  list(
    report = tibble(
      model             = name,
      n_draws           = n_draws,
      max_rhat          = round(max(s$rhat, na.rm = TRUE), 4),
      n_rhat_gt_1.01    = sum(s$rhat > 1.01, na.rm = TRUE),
      min_ess_bulk      = round(min(s$ess_bulk, na.rm = TRUE)),
      min_ess_tail      = round(min(s$ess_tail, na.rm = TRUE)),
      divergences       = div,
      div_pct           = round(100 * div / n_draws, 3),
      pct_max_treedepth = round(100 * mean(td >= 10), 2),
      min_bfmi          = round(min(bfmi, na.rm = TRUE), 3)
    ),
    worst = s |> arrange(ess_bulk) |> head(5)
  )
}

rep1 <- mcmc_report(m1, "M1 linear")
print(as.data.frame(rep1$report))
cat("\nworst-mixing parameters:\n")
print(as.data.frame(rep1$worst))
```

```{r}
#| label: trace
#| fig-height: 6
mcmc_trace(m1, regex_pars = "^b_|^phi$")
```

```{r}
#| label: energy
mcmc_nuts_energy(nuts_params(m1))
```

## Posterior predictive check

```{r}
#| label: ppc
pp_check(m1, ndraws = 100) + labs(title = "PPC — whiff counts, M1")
```

```{r}
#| label: ppc-rate
# The PPC above is on the count scale, where differences in `n_swings` dominate.
# Re-express it as a whiff *rate* so it is comparable with the EDA histogram.
yrep <- posterior_predict(m1, ndraws = 150)
rate_rep <- sweep(yrep, 2, model_df$n_swings, "/")
ppc_dens_overlay(model_df$whiff_rate, rate_rep) +
  labs(title = "PPC on the rate scale — M1")
```

# Testing the linearity assumption

A beta-binomial GLM does not assume the whiff *rate* is linear in mechanics — it
assumes the **logit** of the rate is linear in them. That is the assumption to
test, and the right tool is the residual, not the raw scatter.

**Binned residual plots.** Raw residuals from a binomial-type model are useless
to eyeball, so residuals are averaged within bins of each predictor (Gelman &
Hill). Under a correct specification the binned means scatter around zero with
no trend; roughly 95% should fall inside the $\pm 2\,\text{SE}$ band.

**A weighted RESET-style curvature test.** For each predictor $x$, regress the
residual on $x$ and $x^2$, weighted by number of swings. A non-null quadratic
coefficient is direct evidence of curvature the linear term is not capturing.
A global version regresses the residual on the fitted linear predictor and its
powers. Bonferroni-corrected across predictors.

```{r}
#| label: resid
mu1  <- colMeans(posterior_linpred(m1, transform = TRUE))
eta1 <- colMeans(posterior_linpred(m1))
y_prop <- model_df$n_whiffs / model_df$n_swings
resid1 <- y_prop - mu1
w <- model_df$n_swings

cat(sprintf("weighted mean residual: %+.5f (should be ~0)\n",
            weighted.mean(resid1, w)))
```

```{r}
#| label: binned-resid
#| fig-height: 6
binned_residuals <- function(x, resid, n_bins = 40) {
  o <- order(x)
  tibble(x = x[o], r = resid[o]) |>
    mutate(bin = cut(row_number(), breaks = n_bins, labels = FALSE)) |>
    group_by(bin) |>
    summarise(x = mean(x), resid = mean(r),
              se2 = 2 * sd(r) / sqrt(n()), n = n(), .groups = "drop") |>
    filter(n >= 2)
}

binned_all <- map_dfr(PRED_VARS, function(v) {
  binned_residuals(model_df[[v]], resid1) |> mutate(term = sub("^avg_", "", v))
})

ggplot(binned_all, aes(x = x, y = resid)) +
  geom_ribbon(aes(ymin = -se2, ymax = se2), fill = "steelblue", alpha = 0.15) +
  geom_hline(yintercept = 0) +
  geom_point(size = 1.2, color = "steelblue") +
  geom_smooth(method = "loess", formula = y ~ x, span = 0.75,
              se = FALSE, color = "crimson", linewidth = 0.7) +
  facet_wrap(~ term, scales = "free_x") +
  labs(x = "z-score", y = "mean residual",
       title = "Binned residuals vs predictors — M1 (linear)") +
  theme_minimal()
```

```{r}
#| label: linearity-test
curvature_tests <- function(df, preds, resid, weights) {
  map_dfr(preds, function(v) {
    x <- df[[v]]
    co <- summary(lm(resid ~ x + I(x^2), weights = weights))$coefficients
    tibble(term = v, quad_coef = co[3, 1], std_err = co[3, 2],
           t = co[3, 3], p_value = co[3, 4])
  }) |>
    mutate(p_bonferroni = pmin(1, p_value * length(preds)),
           nonlinear = p_bonferroni < 0.05) |>
    arrange(p_value)
}

curv <- curvature_tests(model_df, PRED_VARS, resid1, w)
print(as.data.frame(curv))

# Global Ramsey RESET on the linear predictor.
reset_null <- lm(resid1 ~ eta1, weights = w)
reset_full <- lm(resid1 ~ eta1 + I(eta1^2) + I(eta1^3), weights = w)
reset_p <- anova(reset_null, reset_full)[["Pr(>F)"]][2]
cat(sprintf("\nglobal RESET (eta^2, eta^3) p-value: %.4g\n", reset_p))

LINEARITY_OK <- !any(curv$nonlinear) && reset_p >= 0.05
cat(sprintf("\nlinearity assumption satisfied: %s\n", LINEARITY_OK))
```

## Follow-up EDA on the non-linear terms

Where the curvature test fires, look directly at the empirical logit — the
observed response on the model's own scale — against the predictor. The
half-continuity correction keeps groups with zero or all whiffs finite.

```{r}
#| label: flagged
# If the tests reject linearity, examine exactly the terms they rejected. If they
# do not, still carry the three most-curved terms forward: Model 2 is the
# nonlinear comparison candidate either way, and a spline model that fails to win
# is itself the evidence that the linear form is adequate.
if (any(curv$nonlinear)) {
  FLAGGED <- curv$term[curv$nonlinear]
  cat("linearity rejected for:", paste(FLAGGED, collapse = ", "), "\n")
} else {
  FLAGGED <- head(curv$term, 3)
  cat("linearity not rejected; carrying most-curved terms as a check:",
      paste(FLAGGED, collapse = ", "), "\n")
}
```

```{r}
#| label: emp-logit
model_df <- model_df |>
  mutate(emp_logit = log((n_whiffs + 0.5) / (n_swings - n_whiffs + 0.5)))

model_df |>
  select(all_of(c(FLAGGED, "emp_logit"))) |>
  pivot_longer(all_of(FLAGGED), names_to = "term", values_to = "z") |>
  ggplot(aes(z, emp_logit)) +
  geom_point(alpha = 0.15, size = 0.6, color = "grey40") +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE,
              color = "crimson", linewidth = 0.9) +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE,
              color = "navy", linetype = "dashed", linewidth = 0.8) +
  facet_wrap(~ term, scales = "free_x") +
  labs(x = "z-score", y = "empirical logit(whiff rate)",
       title = "Empirical logit vs flagged predictors",
       subtitle = "red = LOESS, navy dashed = linear fit") +
  theme_minimal()
```

```{r}
#| label: emp-logit-bytype
# The same curve, split by pitch type: a bend that appears only in the pooled
# view can be Simpson's paradox from differing pitch-type means rather than a
# genuine within-type non-linearity. If it survives the split, it is real.
v <- FLAGGED[1]
top_types <- names(sort(table(model_df$pitch_name), decreasing = TRUE))[1:6]

model_df |>
  filter(pitch_name %in% top_types) |>
  ggplot(aes(.data[[v]], emp_logit, color = pitch_name)) +
  geom_smooth(method = "loess", formula = y ~ x, se = FALSE, linewidth = 0.9) +
  labs(x = paste(v, "(z-score)"), y = "empirical logit(whiff rate)",
       title = paste("Within-pitch-type shape of", v), color = NULL) +
  theme_minimal()
```

# Model 2 — hierarchical spline beta-binomial

Where linearity fails, the fix is to keep the hierarchy and the beta-binomial
likelihood exactly as they are and relax only the *functional form*: replace the
flagged linear terms with a smooth. `s()` fits a **penalized** thin-plate
regression spline, with the wiggliness controlled by a smoothing standard
deviation that Stan estimates alongside everything else. That is preferable to a
fixed-`df` B-spline basis here: an unpenalized basis has to be given exactly the
right flexibility up front, and if it is given too much it will chase noise
between knots, whereas the penalty shrinks unsupported wiggliness toward a
straight line on its own.

The group-specific terms stay linear. Per-pitch-type smooths would add a
smoothing parameter and basis per pitch type, and with a few dozen pitchers in
the smaller types there is not enough information to identify them.

```{r}
#| label: m2-formula
SPLINE_K <- 5

spline_terms <- if_else(PRED_VARS %in% FLAGGED,
                        sprintf("s(%s, k = %d)", PRED_VARS, SPLINE_K),
                        PRED_VARS)

f2 <- as.formula(sprintf(
  "n_whiffs | trials(n_swings) ~ %s + (1 + %s || pitch_name)",
  paste(spline_terms, collapse = " + "),
  paste(PRED_VARS, collapse = " + ")
))
print(f2)

# `sds` is the prior on the smooths' standard deviations -- the penalty. An
# exponential(1) keeps the smooths shrinking toward linear unless the data
# genuinely support curvature.
priors2 <- c(
  prior(normal(0, 1.5), class = "Intercept"),
  prior(normal(0, 1),   class = "b"),
  prior(exponential(1), class = "sd"),
  prior(exponential(1), class = "sds"),
  prior(gamma(0.01, 0.01), class = "phi")
)
```

```{r}
#| label: m2-fit
#| cache: true
m2 <- brm(
  f2, data = model_df, family = beta_binomial(),
  prior = priors2,
  chains = CHAINS, iter = ITER, warmup = WARMUP,
  control = list(adapt_delta = ADAPT_DELTA),
  seed = RANDOM_SEED, refresh = 0
)
summary(m2)
```

```{r}
#| label: m2-diagnostics
rep2 <- mcmc_report(m2, "M2 spline")
print(as.data.frame(bind_rows(rep1$report, rep2$report)))
cat("\nworst-mixing parameters (M2):\n")
print(as.data.frame(rep2$worst))
```

```{r}
#| label: m2-smooths
#| fig-height: 3.2
# What the penalty actually decided: a smooth shrunk to a straight line is the
# model telling you the linear term was adequate.
plot(conditional_smooths(m2), ask = FALSE)
```

```{r}
#| label: m2-resid
#| fig-height: 3.2
mu2 <- colMeans(posterior_linpred(m2, transform = TRUE))
resid2 <- y_prop - mu2

cat("curvature tests after splines:\n")
print(as.data.frame(curvature_tests(model_df, PRED_VARS, resid2, w)))

map_dfr(FLAGGED, function(v) {
  binned_residuals(model_df[[v]], resid2) |> mutate(term = sub("^avg_", "", v))
}) |>
  ggplot(aes(x, resid)) +
  geom_ribbon(aes(ymin = -se2, ymax = se2), fill = "seagreen", alpha = 0.15) +
  geom_hline(yintercept = 0) +
  geom_point(size = 1.2, color = "seagreen") +
  geom_smooth(method = "loess", formula = y ~ x, span = 0.75,
              se = FALSE, color = "crimson", linewidth = 0.7) +
  facet_wrap(~ term, scales = "free_x") +
  labs(x = "z-score", y = "mean residual",
       title = "Binned residuals — M2 (penalized splines on flagged terms)") +
  theme_minimal()
```

## Fit comparison by LOO

```{r}
#| label: loo
loo1 <- loo(m1)
loo2 <- loo(m2)
comparison <- loo_compare(list("M1 linear" = loo1, "M2 spline" = loo2))
print(comparison)
```

PSIS-LOO is an importance-sampling approximation and it can fail silently. Two
quantities say whether the numbers above mean anything: the Pareto shape $k$ per
observation (above ~0.7 the importance weights have infinite variance) and
`p_loo`, the effective number of parameters. `p_loo` exceeding the number of
observations is not a large model — it is a broken estimate. When that happens
the fix is *not* to tune the model until the diagnostic stops complaining, which
would be selecting on the diagnostic rather than on predictive accuracy; it is
to fall back on cross-validation, which refits and needs no importance weights.
Nothing here depends on it: the selection criterion below is out-of-sample
$R^2$ / MSE on run value, and that is a $K$-fold estimate already.

```{r}
#| label: loo-reliability
loo_reliability <- function(l, name) {
  k <- l$diagnostics$pareto_k
  n <- length(k)
  bad <- sum(k > 0.7)
  p_loo <- l$estimates["p_loo", "Estimate"]
  reliable <- (p_loo < n / 2) && (bad / n <= 0.05)
  cat(sprintf("%-10s elpd=%10.1f  p_loo=%9.1f  (n=%d)  k>0.7: %4d (%4.1f%%)  max k=%.2f  -> %s\n",
              name, l$estimates["elpd_loo", "Estimate"], p_loo, n,
              bad, 100 * bad / n, max(k), if (reliable) "ok" else "UNRELIABLE"))
  reliable
}

LOO_RELIABLE <- c(
  "M1 linear" = loo_reliability(loo1, "M1 linear"),
  "M2 spline" = loo_reliability(loo2, "M2 spline")
)
LOO_TRUSTWORTHY <- all(LOO_RELIABLE)
if (!LOO_TRUSTWORTHY) {
  cat("\nLOO cannot rank these models and the elpd column should not be quoted.",
      "\nThe comparison that does not depend on importance weights is the",
      "\ngrouped K-fold run-value evaluation below, which is the stated",
      "\nselection criterion anyway.\n")
}
```

# Per-pitch-type posteriors

The hierarchy gives every pitch type its own posterior. Two views of it: the
posterior for each type's *slope* on a mechanic (does velocity buy more whiffs
on a four-seamer than on a changeup?), and the posterior for each type's whiff
rate at that type's own average mechanics.

```{r}
#| label: forest-slopes
FOCUS <- "avg_release_speed"

# The group-level deviations r_pitch_name[type, term] are offsets from the
# population slope b_<term>; adding them back gives each type's own slope.
slope_draws <- as_draws_df(m2) |>
  select(matches(sprintf("^b_%s$|^r_pitch_name\\[.*,%s\\]$", FOCUS, FOCUS)))
pop <- slope_draws[[paste0("b_", FOCUS)]]

type_slopes <- slope_draws |>
  select(-all_of(paste0("b_", FOCUS))) |>
  imap_dfr(~ tibble(
    pitch_name = sub(sprintf("^r_pitch_name\\[(.*),%s\\]$", FOCUS), "\\1", .y),
    slope = .x + pop
  ))

type_slopes |>
  group_by(pitch_name) |>
  summarise(mean = mean(slope),
            lo = quantile(slope, 0.03), hi = quantile(slope, 0.97),
            .groups = "drop") |>
  bind_rows(tibble(pitch_name = "POPULATION", mean = mean(pop),
                   lo = quantile(pop, 0.03), hi = quantile(pop, 0.97))) |>
  mutate(pitch_name = fct_reorder(pitch_name, mean)) |>
  ggplot(aes(mean, pitch_name)) +
  geom_vline(xintercept = 0, linetype = "dotted") +
  geom_pointrange(aes(xmin = lo, xmax = hi)) +
  labs(x = sprintf("posterior slope on %s (log-odds per SD)", FOCUS),
       y = NULL, title = "Population slope and per-pitch-type slopes") +
  theme_minimal()
```

```{r}
#| label: type-posterior
type_posterior <- function(fit, df, preds) {
  grid <- df |>
    group_by(pitch_name) |>
    summarise(across(all_of(preds), mean), .groups = "drop") |>
    mutate(n_swings = 1)          # trials must be present; transform gives mu
  mu <- posterior_linpred(fit, newdata = grid, transform = TRUE)
  colnames(mu) <- as.character(grid$pitch_name)
  list(grid = grid, mu = mu)
}

tp <- type_posterior(m2, model_df, PRED_VARS)

summary_by_type <- tibble(
  pitch_name = colnames(tp$mu),
  mean = colMeans(tp$mu),
  sd   = apply(tp$mu, 2, sd),
  lo   = apply(tp$mu, 2, quantile, 0.03),
  hi   = apply(tp$mu, 2, quantile, 0.97)
) |> arrange(desc(mean))
print(as.data.frame(summary_by_type |> mutate(across(where(is.numeric), ~ round(.x, 3)))))

summary_by_type |>
  mutate(pitch_name = fct_reorder(pitch_name, mean)) |>
  ggplot(aes(mean, pitch_name)) +
  geom_pointrange(aes(xmin = lo, xmax = hi)) +
  labs(x = "posterior whiff rate at pitch-type-average mechanics",
       y = NULL, title = "Per-pitch-type posterior") +
  theme_minimal()
```

# Stuff+

**Definition.** For posterior draw $d$, pitcher $i$ throwing pitch type $k$, let
$\eta_{id}$ be the model's latent log-odds of a whiff. Standardize it against the
posterior for *that pitch type*:

$$
z_{id} \;=\; \frac{\eta_{id} - m_{kd}}{s_{kd}},
\qquad
\text{Stuff+}_{id} \;=\; 100 + 10\,z_{id}
$$

where $m_{kd}$ and $s_{kd}$ are the mean and standard deviation of $\eta_{\cdot d}$
across the pitchers who throw type $k$, computed **within each draw**. So 100 is
average for that pitch type, one standard deviation is 10 points, and higher is
better. Standardizing draw-by-draw rather than on posterior means lets Stuff+
inherit the model's uncertainty: each pitcher gets a posterior *distribution* of
Stuff+, not a point estimate.

```{r}
#| label: stuffplus
stuff_plus_draws <- function(fit, df, center = 100, scale = 10) {
  eta <- posterior_linpred(fit)          # draws x observations, link scale
  codes <- as.integer(df$pitch_name)
  z <- eta
  for (k in unique(codes)) {
    sel <- which(codes == k)
    sub <- eta[, sel, drop = FALSE]
    z[, sel] <- (sub - rowMeans(sub)) / apply(sub, 1, sd)
  }
  center + scale * z
}

# Highest-density interval: the shortest interval containing `prob` of the mass.
hdi1 <- function(x, prob = 0.94) {
  x <- sort(x); n <- length(x); m <- max(1L, floor(prob * n))
  i <- which.min(x[(m + 1):n] - x[1:(n - m)])
  c(x[i], x[i + m])
}

summarise_stuff <- function(sp, df) {
  bounds <- t(apply(sp, 2, hdi1))
  df |> mutate(
    stuff_plus    = colMeans(sp),
    stuff_plus_sd = apply(sp, 2, sd),
    stuff_plus_lo = bounds[, 1],
    stuff_plus_hi = bounds[, 2]
  )
}

stuff1 <- summarise_stuff(stuff_plus_draws(m1, model_df), model_df)
stuff2 <- summarise_stuff(stuff_plus_draws(m2, model_df), model_df)

stuff1 |>
  group_by(pitch_name) |>
  summarise(mean = mean(stuff_plus), sd = sd(stuff_plus),
            min = min(stuff_plus), max = max(stuff_plus), .groups = "drop") |>
  mutate(across(where(is.numeric), ~ round(.x, 2))) |>
  as.data.frame() |>
  print()
```

```{r}
#| label: stuff-leaders
show_cols <- c("player_name", "pitch_name", "n_pitches", "whiff_rate",
               "stuff_plus", "stuff_plus_sd", "stuff_plus_lo", "stuff_plus_hi",
               "run_value")
cat("top 15 by Stuff+ (M1)\n")
stuff1 |> slice_max(stuff_plus, n = 15) |> select(all_of(show_cols)) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |> as.data.frame() |> print()
cat("\nbottom 10 by Stuff+ (M1)\n")
stuff1 |> slice_min(stuff_plus, n = 10) |> select(all_of(show_cols)) |>
  mutate(across(where(is.numeric), ~ round(.x, 3))) |> as.data.frame() |> print()
```

```{r}
#| label: stuff-uncertainty
p_dist <- ggplot(stuff1, aes(stuff_plus)) +
  geom_histogram(bins = 40, fill = "steelblue", color = "white") +
  geom_vline(xintercept = 100, color = "red", linetype = "dashed") +
  labs(x = "Stuff+", y = "count", title = "Stuff+ distribution (M1)") +
  theme_minimal()

p_unc <- ggplot(stuff1, aes(n_pitches, stuff_plus_sd)) +
  geom_point(alpha = 0.35, size = 0.9) +
  scale_x_log10() +
  labs(x = "pitches thrown (log)", y = "posterior SD of Stuff+",
       title = "Uncertainty shrinks with sample size") +
  theme_minimal()

print(p_dist); print(p_unc)
```

# Model selection on cumulative run value

The criterion is **$R^2$ and MSE for cumulative run value**. Two things have to
be handled for that comparison to mean anything.

*Exposure.* `run_value` is a **sum** over pitches, so it scales with usage: a
pitcher who threw 600 sliders has a larger $|{\rm run\ value}|$ than one who
threw 40, regardless of quality. Any evaluation that ignores `n_pitches` mostly
measures playing time. Pitch count is therefore in every specification,
including the baseline, so the reported gain is Stuff+'s **incremental**
contribution rather than a restatement of usage.

*Leakage.* Rows are pitcher × pitch-type, and one pitcher contributes several
rows. Splitting rows at random would put the same pitcher's talent on both sides
of the fold, so cross-validation is grouped by pitcher.

```{r}
#| label: eval
TARGET   <- "run_value"      # cumulative; negative favours the pitcher
N_SPLITS <- 5
BASELINE <- "baseline (exposure + pitch type)"

group_kfold <- function(groups, k = N_SPLITS, seed = RANDOM_SEED) {
  set.seed(seed)
  u <- unique(groups)
  assign <- setNames(sample(rep_len(seq_len(k), length(u))), u)
  unname(assign[as.character(groups)])
}

r2 <- function(y, pred) 1 - sum((y - pred)^2) / sum((y - mean(y))^2)

evaluate <- function(df, label, use_stuff = TRUE, target = TARGET) {
  rhs <- c("n_pitches", "pitch_name")
  if (use_stuff) {
    # Stuff+ is standardized within pitch type, so its effect on a *cumulative*
    # total must scale with how often the pitch was thrown.
    rhs <- c(rhs, "stuff_plus", "stuff_plus:n_pitches")
  }
  fml <- reformulate(rhs, response = target)

  folds <- group_kfold(df$player_name)
  oof <- numeric(nrow(df))
  for (i in seq_len(N_SPLITS)) {
    te <- folds == i
    fit <- lm(fml, data = df[!te, ])
    oof[te] <- predict(fit, newdata = df[te, ])
  }
  ins <- predict(lm(fml, data = df))
  y <- df[[target]]

  tibble(model = label,
         cv_r2 = r2(y, oof), cv_mse = mean((y - oof)^2),
         cv_rmse = sqrt(mean((y - oof)^2)),
         in_sample_r2 = r2(y, ins), in_sample_mse = mean((y - ins)^2))
}

results <- bind_rows(
  evaluate(stuff1, BASELINE, use_stuff = FALSE),
  evaluate(stuff1, "M1 linear -> Stuff+"),
  evaluate(stuff2, "M2 spline -> Stuff+")
)
cat("target: cumulative run value\n")
print(as.data.frame(results |> mutate(across(where(is.numeric), ~ round(.x, 4)))))
```

```{r}
#| label: eval-rate
# Secondary view on the rate scale, where exposure is already divided out.
results_rate <- bind_rows(
  evaluate(stuff1, BASELINE, use_stuff = FALSE, target = "rv_per_100"),
  evaluate(stuff1, "M1 linear -> Stuff+", target = "rv_per_100"),
  evaluate(stuff2, "M2 spline -> Stuff+", target = "rv_per_100")
)
cat("target: run value per 100 pitches\n")
print(as.data.frame(results_rate |> mutate(across(where(is.numeric), ~ round(.x, 4)))))
```

```{r}
#| label: select
# For a fixed target, R^2 = 1 - MSE / Var(y). The two criteria are monotone
# transforms of one another and therefore always produce the same ranking --
# reporting both is still worthwhile because MSE is on the scale of runs^2 and
# says how large the errors actually are, but it cannot break a tie. The real
# question is whether the leader's margin justifies the extra parameters.
candidates <- results |> filter(model != BASELINE) |> arrange(desc(cv_r2))
print(as.data.frame(candidates |> select(model, cv_r2, cv_mse, cv_rmse) |>
                      mutate(across(where(is.numeric), ~ round(.x, 4)))))
stopifnot(candidates$model[1] == results |> filter(model != BASELINE) |>
            slice_min(cv_mse, n = 1) |> pull(model))

PARSIMONY_TOL <- 0.005
leader <- candidates$model[1]
margin <- candidates$cv_r2[1] - candidates$cv_r2[2]

if (startsWith(leader, "M2") && margin < PARSIMONY_TOL) {
  SELECTED <- candidates$model[2]
  cat(sprintf("\n%s leads by only %.4f R^2, inside the %.3f tolerance -- keeping the simpler linear model.\n",
              leader, margin, PARSIMONY_TOL))
} else {
  SELECTED <- leader
  cat(sprintf("\n%s leads by %.4f R^2.\n", leader, margin))
}
cat("SELECTED:", SELECTED, "\n")

if (LOO_TRUSTWORTHY) {
  cat("LOO preference (whiff fit, for reference):", rownames(comparison)[1], "\n")
} else {
  cat("LOO withheld: unreliable for",
      paste(names(LOO_RELIABLE)[!LOO_RELIABLE], collapse = ", "),
      "(see the Pareto-k check)\n")
}

best_stuff <- if (startsWith(SELECTED, "M1")) stuff1 else stuff2
gain <- candidates$cv_r2[candidates$model == SELECTED] -
  results$cv_r2[results$model == BASELINE]
cat(sprintf("incremental out-of-sample R^2 from Stuff+: %+.4f (baseline %.4f -> %.4f)\n",
            gain, results$cv_r2[results$model == BASELINE],
            candidates$cv_r2[candidates$model == SELECTED]))
```

```{r}
#| label: eval-plot
ggplot(best_stuff, aes(stuff_plus, rv_per_100)) +
  geom_point(aes(color = n_pitches), alpha = 0.6, size = 1.2) +
  scale_color_viridis_c(trans = "log10", name = "pitches") +
  geom_smooth(method = "lm", formula = y ~ x, se = FALSE, color = "crimson") +
  geom_hline(yintercept = 0, linewidth = 0.3) +
  geom_vline(xintercept = 100, linetype = "dotted", linewidth = 0.3) +
  annotate("text", x = -Inf, y = -Inf, hjust = -0.2, vjust = -1,
           label = sprintf("r = %.3f",
                           cor(best_stuff$stuff_plus, best_stuff$rv_per_100))) +
  labs(x = "Stuff+", y = "run value per 100 pitches",
       title = paste("Stuff+ vs run value —", SELECTED)) +
  theme_minimal()
```

```{r}
#| label: export
out_cols <- c("player_name", "pitch_name", "n_pitches", "n_swings", "n_whiffs",
              "whiff_rate", "run_value", "rv_per_100",
              "stuff_plus", "stuff_plus_sd", "stuff_plus_lo", "stuff_plus_hi")

best_stuff |>
  arrange(pitch_name, desc(stuff_plus)) |>
  select(all_of(out_cols)) |>
  write_csv(file.path(DATA_DIR, "stuff_plus.csv"))

bind_rows(
  results      |> mutate(target = "cumulative_run_value"),
  results_rate |> mutate(target = "run_value_per_100")
) |>
  relocate(target) |>
  write_csv(file.path(DATA_DIR, "model_selection.csv"))

cat("wrote", file.path(DATA_DIR, "stuff_plus.csv"), "and",
    file.path(DATA_DIR, "model_selection.csv"), "\n")
cat("selected model:", SELECTED, "\n")
```

# Summary

| Stage | Result |
|---|---|
| Missing data | `arm_angle` is absent for some pitchers entirely; incomplete rows dropped explicitly before standardizing |
| Collinearity | VIF measured *within* pitch type (the varying intercept absorbs between-type differences), pruned at 5 |
| Overdispersion | `beta_binomial()` over `binomial()`, `phi` estimated |
| Hierarchy | Varying intercept + uncorrelated varying slopes by `pitch_name`; each type gets its own posterior |
| Linearity | Binned residuals + weighted curvature tests + global RESET; penalized `s()` smooths on the terms that fail |
| MCMC | $\hat{R}$, ESS bulk/tail, divergences, tree depth, E-BFMI, energy plot, PPC on count and rate scales |
| Fit comparison | LOO, gated on a Pareto-$k$ reliability check and withheld when PSIS degenerates |
| Stuff+ | Draw-wise z-score against the pitch type's own posterior, 100 ± 10 |
| Selection | Grouped 5-fold CV $R^2$ / MSE on cumulative run value, with exposure in the baseline |
