---
title: "MLB Statcast: A Hierarchical Bayesian Stuff+ Model"
author: "Trey Chase"
format:
  pdf:
    toc: true
    number-sections: true
    fig-width: 7
    fig-height: 4.5
editor: visual
---

# Overview

This notebook builds a **Stuff+** metric for individual pitches from Statcast
data. The structure is:

1.  Scrape pitch-level Statcast data and reduce it to a pitcher × pitch-type
    table of average release/movement mechanics, whiff rate, and **cumulative
    run value**.
2.  Fit a **hierarchical Bayesian beta-binomial model** of whiff propensity on
    mechanics, with a varying intercept *and* varying slopes by `pitch_name`.
    Each pitch type therefore gets its own posterior.
3.  Define **Stuff+** as the standardized z-score of a pitcher's posterior
    latent whiff propensity relative to the posterior for *his own pitch type*,
    on the conventional 100 ± 10 scale.
4.  **Test the linearity assumption** of the logit-linear predictor. Where it
    fails, do targeted EDA and fit a second MCMC model with spline bases.
5.  Run full MCMC diagnostics, then **select the model by out-of-sample
    $R^2$ and MSE against cumulative run value**.

```{r}
#| label: setup
#| message: false
#| warning: false

library(baseballr)
library(tidyverse)
library(purrr)

knitr::opts_chunk$set(message = FALSE, warning = FALSE)

# Everything is written relative to the project directory so that the R and
# Python halves of this notebook read and write the *same* file. (Previously R
# wrote pitcher_summary.csv to the working directory while Python read a copy
# from ~/Downloads, so any re-run silently modelled a stale export.)
DATA_DIR <- "data"
dir.create(DATA_DIR, showWarnings = FALSE, recursive = TRUE)
SUMMARY_PATH <- file.path(DATA_DIR, "pitcher_summary.csv")
```

# Data acquisition

```{r}
#| label: scrape
#| cache: true

date_starts <- seq(as.Date("2026-04-10"), Sys.Date(), by = "7 days")
date_ends   <- pmin(date_starts + 6, Sys.Date())

data <- map2_dfr(
  date_starts, date_ends,
  ~ scrape_statcast_savant(
      start_date  = .x,
      end_date    = .y,
      player_type = "pitcher"
    )
)

nrow(data)
```

# Pitch-level cleaning

Three changes relative to a naive `select() |> distinct()`:

**Run value is carried through.** `delta_run_exp` is Statcast's change in run
expectancy attributable to each pitch (negative = good for the pitcher). It is
the outcome we ultimately validate against, so it has to survive the reduction.

**De-duplication uses the true pitch key.** `distinct()` over the *value*
columns silently collapses two genuinely different pitches that happen to share
rounded measurements. `game_pk` + `at_bat_number` + `pitch_number` uniquely
identifies a pitch, so dedupe on that instead.

**Horizontal quantities are mirrored for left-handers.** `pfx_x` and
`release_pos_x` are recorded in the catcher's frame, so a lefty's arm-side run
carries the opposite sign to a righty's. Averaging the two together — as the
original pipeline did — pushes group means toward zero and makes the coefficient
on `avg_pfx_x` close to meaningless. Flipping the sign for LHP puts every pitch
in a **pitcher-centric** frame: positive `pfx_x` is arm-side, negative is glove-side.

```{r}
#| label: pitch-df

# `arm_angle` was added to the Statcast feed relatively recently; older
# baseballr versions do not return it and `select()` would abort. Create it as
# missing instead and let the Python side's missingness report and VIF screen
# decide what to do with it. `delta_run_exp` is not optional -- it is the
# model-selection target -- so fail loudly if it is absent.
if (!"arm_angle" %in% names(data)) {
  warning("arm_angle not returned by this baseballr version; filling with NA")
  data$arm_angle <- NA_real_
}
stopifnot("delta_run_exp missing - update baseballr" =
            "delta_run_exp" %in% names(data))

pitch_df <- data |>
  select(
    game_pk, at_bat_number, pitch_number,
    player_name, pitcher, p_throws, stand,
    pitch_name,
    release_speed, release_pos_x, release_pos_z,
    release_spin_rate, release_extension, spin_axis, arm_angle,
    pfx_x, pfx_z, plate_x, plate_z,
    description, delta_run_exp
  ) |>
  distinct(game_pk, at_bat_number, pitch_number, .keep_all = TRUE)

# Drop pitch types that are too rare or too heterogeneous to model.
pitch_df <- pitch_df |>
  filter(!pitch_name %in% c(
    "Forkball", "Pitch Out", "Eephus", "Knuckleball",
    "Other", "Slow Curve", "Unknown", "NA"
  ), !is.na(pitch_name), pitch_name != "")

# Pitcher-centric horizontal frame.
pitch_df <- pitch_df |>
  mutate(
    pfx_x         = if_else(p_throws == "L", -pfx_x, pfx_x),
    release_pos_x = if_else(p_throws == "L", -release_pos_x, release_pos_x),
    spin_axis     = if_else(p_throws == "L", (360 - spin_axis) %% 360, spin_axis)
  )

nrow(pitch_df)
```

# Exploratory data analysis

```{r}
#| label: eda-velo
ggplot(pitch_df, aes(x = release_speed, fill = pitch_name)) +
  geom_histogram(bins = 30, color = "white") +
  facet_wrap(~ pitch_name, scales = "free") +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = "Release Speed (mph)", y = "Count",
       title = "Release Speed Distribution by Pitch Type")
```

```{r}
#| label: eda-release
ggplot(pitch_df, aes(x = release_pos_x, y = release_pos_z, color = p_throws)) +
  geom_point(alpha = 0.25, size = 0.5) +
  labs(
    x = "Release Position X (ft, arm-side positive)",
    y = "Release Position Z (ft)",
    title = "Release Position by Pitcher Handedness (mirrored for LHP)",
    color = "Throws"
  ) +
  theme_minimal()
```

After mirroring, the two handedness clouds overlap instead of sitting on
opposite sides of the plate — which is the point: a left-hander's slider and a
right-hander's slider now occupy the same region of mechanics space and can
share a pitch-type-level prior.

```{r}
#| label: eda-movement
ggplot(pitch_df, aes(x = pfx_x * 12, y = pfx_z * 12, color = pitch_name)) +
  geom_point(alpha = 0.15, size = 0.4) +
  coord_fixed() +
  labs(x = "Horizontal break (in, arm-side positive)",
       y = "Induced vertical break (in)",
       title = "Movement Profile by Pitch Type", color = NULL) +
  theme_minimal()
```

# Aggregation to pitcher × pitch type

```{r}
#| label: summarise

swing_descriptions <- c(
  "foul", "foul_tip", "hit_into_play", "swinging_strike",
  "swinging_strike_blocked", "missed_bunt", "foul_bunt"
)
whiff_descriptions <- c("swinging_strike", "swinging_strike_blocked", "missed_bunt")

pitch_df <- pitch_df |>
  mutate(
    is_swing = description %in% swing_descriptions,
    is_whiff = description %in% whiff_descriptions
  )

pitcher_summary <- pitch_df |>
  group_by(pitch_name, player_name) |>
  summarise(
    avg_release_speed     = mean(release_speed, na.rm = TRUE),
    avg_release_pos_x     = mean(release_pos_x, na.rm = TRUE),
    avg_release_pos_z     = mean(release_pos_z, na.rm = TRUE),
    avg_release_spin_rate = mean(release_spin_rate, na.rm = TRUE),
    avg_release_extension = mean(release_extension, na.rm = TRUE),
    avg_arm_angle         = mean(arm_angle, na.rm = TRUE),
    avg_pfx_x             = mean(pfx_x, na.rm = TRUE),
    avg_pfx_z             = mean(pfx_z, na.rm = TRUE),
    n_swings  = sum(is_swing, na.rm = TRUE),
    n_whiffs  = sum(is_whiff, na.rm = TRUE),
    n_pitches = n(),
    # Cumulative run value: the model-selection target. Negative favours the
    # pitcher, so it is negated into `run_value_saved` for readability.
    run_value = sum(delta_run_exp, na.rm = TRUE),
    .groups = "drop"
  ) |>
  filter(n_pitches >= 20, n_swings >= 10) |>
  mutate(
    whiff_rate       = n_whiffs / n_swings,
    rv_per_100       = 100 * run_value / n_pitches,
    run_value_saved  = -run_value
  ) |>
  arrange(pitch_name, player_name)

# `mean(x, na.rm = TRUE)` returns NaN when a group is entirely missing, so a
# group can clear the n_pitches filter and still carry NaN mechanics. This is
# exactly what later produced "'data' contains N incomplete rows" in the Python
# model fit; it is reported here and handled explicitly on the Python side.
mech_cols <- c(
  "avg_release_speed", "avg_release_pos_x", "avg_release_pos_z",
  "avg_release_spin_rate", "avg_release_extension", "avg_arm_angle",
  "avg_pfx_x", "avg_pfx_z"
)

pitcher_summary |>
  summarise(across(all_of(mech_cols), ~ sum(!is.finite(.x)))) |>
  pivot_longer(everything(), names_to = "column", values_to = "n_missing") |>
  arrange(desc(n_missing))
```

```{r}
#| label: whiff-hist
wr_mean <- mean(pitcher_summary$whiff_rate, na.rm = TRUE)
wr_var  <- var(pitcher_summary$whiff_rate, na.rm = TRUE)

ggplot(pitcher_summary, aes(x = whiff_rate)) +
  geom_histogram(bins = 30, fill = "steelblue", color = "white") +
  geom_vline(xintercept = wr_mean, color = "red", linetype = "dashed") +
  annotate("text", x = wr_mean, y = Inf,
           label = paste0("Mean = ", round(wr_mean, 3),
                          "\nVar = ", round(wr_var, 3)),
           vjust = 1.5, hjust = -0.1, color = "red") +
  labs(x = "Whiff Rate (Whiffs / Swings)", y = "Count",
       title = "Whiff Rate by Pitcher and Pitch Type (20+ pitches, 10+ swings)") +
  theme_minimal()
```

The binomial variance implied by the mean whiff rate is far smaller than the
observed variance across pitcher-pitch rows, so the counts are **overdispersed**
relative to a binomial. That is the substantive reason for the `beta_binomial`
family below rather than plain `binomial`.

```{r}
#| label: rv-hist
ggplot(pitcher_summary, aes(x = run_value)) +
  geom_histogram(bins = 40, fill = "darkorange", color = "white") +
  labs(x = "Cumulative run value (runs, negative favours pitcher)",
       y = "Count", title = "Cumulative Run Value by Pitcher and Pitch Type") +
  theme_minimal()
```

```{r}
#| label: export
write_csv(pitcher_summary, SUMMARY_PATH)
SUMMARY_PATH
```

# Bayesian model: whiff propensity ~ pitch mechanics

```{python}
#| label: py-setup
import warnings

import arviz as az
import bambi as bmb
import matplotlib.pyplot as plt
import numpy as np
import pandas as pd
from scipy.special import logit

warnings.filterwarnings("ignore", category=FutureWarning)
RANDOM_SEED = 42
rng = np.random.default_rng(RANDOM_SEED)

# nutpie is a drop-in NUTS implementation that compiles the model; it is several
# times faster than the default PyMC sampler on models this size. Fall back
# silently if it is not installed.
try:
    import nutpie  # noqa: F401
    NUTS_SAMPLER = "nutpie"
except ModuleNotFoundError:
    NUTS_SAMPLER = "pymc"

SAMPLE_KW = dict(
    draws=2000,
    tune=2000,
    chains=4,
    target_accept=0.95,
    random_seed=RANDOM_SEED,
    # bambi selects the backend with `inference_method=`; passing pm.sample's
    # own `nuts_sampler=` straight through raises "got multiple values for
    # keyword argument 'nuts_sampler'".
    inference_method=NUTS_SAMPLER,
    idata_kwargs={"log_likelihood": True},
    # bambi's default `omit_offsets=True` drops the non-centered `*_offset`
    # variables from the posterior. Those are exactly what the model needs to
    # reconstruct its group-specific terms, so computing the log-likelihood
    # afterwards fails with KeyError: '<term>|pitch_name_offset'. Keeping them
    # costs a little memory and makes LOO possible.
    omit_offsets=False,
    progressbar=False,
)
print(f"sampler: {NUTS_SAMPLER}")


def ensure_log_likelihood(model, idata):
    """LOO needs a pointwise log-likelihood. nutpie silently ignores
    `idata_kwargs={"log_likelihood": True}` -- it warns and moves on -- so
    compute it after the fact when it is missing. Cheap next to sampling."""
    if "log_likelihood" not in idata.groups():
        import pymc as pm
        missing = [v for v in model.backend.model.free_RVs
                   if v.name not in idata.posterior]
        if missing:
            raise RuntimeError(
                "cannot compute the log-likelihood: "
                f"{[v.name for v in missing]} are absent from the posterior. "
                "Fit with omit_offsets=False."
            )
        with model.backend.model:
            pm.compute_log_likelihood(idata, extend_inferencedata=True,
                                      progressbar=False)
    return idata
```

## Loading and data hygiene

The original fit died here:

```
ValueError: 'data' contains 5 incomplete rows.
```

`formulae` (the design-matrix backend behind `bambi`) refuses to silently drop
missing rows. The missingness is real: `arm_angle` is a relatively new Statcast
field and is absent for some pitchers entirely, and when *every* pitch in a
group is missing it, `mean(..., na.rm = TRUE)` returns `NaN` rather than
dropping the group. The `NameError: name 'model' is not defined` that followed
was purely a cascade — `bmb.Model(...)` raised, so `model` was never bound and
every downstream chunk failed on the same name.

The fix is to drop the incomplete rows **explicitly and visibly**, before
standardizing, so that the means and standard deviations used to z-score the
predictors are computed on exactly the rows that get modelled.

```{python}
#| label: py-load
summary_path = "data/pitcher_summary.csv"
raw = pd.read_csv(summary_path)

MECH_VARS = [
    "avg_release_speed", "avg_release_pos_x", "avg_release_pos_z",
    "avg_release_spin_rate", "avg_release_extension", "avg_arm_angle",
    "avg_pfx_x", "avg_pfx_z",
]

# `mean(x, na.rm = TRUE)` on an all-missing group yields NaN, which read_csv may
# parse as either NaN or +/-inf; treat both as missing.
raw[MECH_VARS] = raw[MECH_VARS].replace([np.inf, -np.inf], np.nan)

missing_report = (
    raw[MECH_VARS].isna().sum()
    .rename("n_missing").to_frame()
    .assign(pct=lambda d: (100 * d.n_missing / len(raw)).round(2))
    .sort_values("n_missing", ascending=False)
)
print(missing_report)

incomplete = raw[MECH_VARS].isna().any(axis=1)
print(f"\n{incomplete.sum()} of {len(raw)} rows incomplete "
      f"({100 * incomplete.mean():.2f}%) -> dropped")
print(raw.loc[incomplete, "pitch_name"].value_counts().to_string())

df = raw.loc[~incomplete].reset_index(drop=True).copy()
```

Because the dropped rows are concentrated in whichever pitch types happen to
lack `arm_angle` coverage rather than being spread at random, this is
missing-not-completely-at-random. At well under 2% of rows it is not worth
imputing, but it *is* worth checking that no pitch type loses a meaningful
share of its pitchers — the group-level terms below need enough pitchers per
type to estimate a variance.

```{python}
#| label: py-group-sizes
sizes = (
    pd.DataFrame({"before": raw.pitch_name.value_counts(),
                  "after": df.pitch_name.value_counts()})
    .fillna(0).astype(int)
    .assign(pct_lost=lambda d: (100 * (1 - d.after / d.before)).round(1))
    .sort_values("after", ascending=False)
)
print(sizes)

# A pitch type with only a handful of pitchers cannot support its own slopes;
# it would be shrunk almost entirely to the population mean anyway, and it
# destabilises the group-level variances. Require a minimum group size.
MIN_PITCHERS_PER_TYPE = 25
keep_types = sizes.index[sizes["after"] >= MIN_PITCHERS_PER_TYPE]
dropped_types = sorted(set(sizes.index) - set(keep_types))
if dropped_types:
    print(f"\ndropping sparse pitch types: {dropped_types}")
df = df[df.pitch_name.isin(keep_types)].reset_index(drop=True)
print(f"modelling rows: {len(df)}, pitch types: {df.pitch_name.nunique()}")
```

## Collinearity screen

Before fitting, check whether the eight mechanics are separately identifiable.
`arm_angle` is a particular concern: Statcast derives it from the release point
and the pitcher's shoulder position, so it is close to a deterministic function
of `release_pos_x` and `release_pos_z`. Slopes on mutually collinear predictors
are individually unstable even when the model's predictions are fine — and
unstable slopes are exactly what Stuff+ is built out of.

```{python}
#| label: py-vif
from statsmodels.stats.outliers_influence import variance_inflation_factor

def within_group_center(frame, cols, group="pitch_name"):
    """Center each predictor on its own pitch-type mean."""
    return frame.groupby(group, observed=True)[cols].transform(lambda s: s - s.mean())

def vif_table(frame, cols):
    X = np.column_stack([np.ones(len(frame))] + [frame[c].to_numpy() for c in cols])
    return (
        pd.Series({c: variance_inflation_factor(X, i + 1) for i, c in enumerate(cols)},
                  name="VIF")
        .sort_values(ascending=False).to_frame().round(2)
    )

# The model already carries a varying intercept for `pitch_name`, so between-type
# differences are absorbed there and do not destabilise the slopes. What matters
# is collinearity *within* a pitch type. Pooled VIF is dominated by the fact that
# sliders are simply slower and spin differently than four-seamers, and would
# throw away genuinely informative movement variables for the wrong reason.
centered = within_group_center(df, MECH_VARS)

print("pooled correlation (between + within pitch type)")
print(df[MECH_VARS].corr().round(2).to_string())
print("\nwithin-pitch-type correlation")
print(centered.corr().round(2).to_string())
print("\npooled VIF (misleading here)")
print(vif_table(df, MECH_VARS).T.to_string())
print("\nwithin-pitch-type VIF (the one that matters)")
print(vif_table(centered, MECH_VARS).T.to_string())
```

```{python}
#| label: py-vif-prune
VIF_THRESHOLD = 5.0

def prune_by_vif(frame, cols, threshold=VIF_THRESHOLD):
    """Iteratively drop the single worst-inflated predictor until all VIF < threshold."""
    cols, dropped = list(cols), []
    while len(cols) > 1:
        v = vif_table(frame, cols)["VIF"]
        if v.max() < threshold:
            break
        worst = v.idxmax()
        dropped.append((worst, round(float(v.max()), 2)))
        cols.remove(worst)
    return cols, dropped

PRED_VARS, vif_dropped = prune_by_vif(centered, MECH_VARS)
for name, val in vif_dropped:
    print(f"dropped {name} (within-type VIF = {val})")
print("\nretained predictors:", PRED_VARS)
print(vif_table(centered, PRED_VARS).T.to_string())
```

## Standardization and response

```{python}
#| label: py-standardize
STD_STATS = df[PRED_VARS].agg(["mean", "std"])
print(STD_STATS.round(3).to_string())

model_df = df.copy()
for v in PRED_VARS:
    model_df[v] = (model_df[v] - STD_STATS.loc["mean", v]) / STD_STATS.loc["std", v]

model_df["pitch_name"] = pd.Categorical(model_df["pitch_name"])
model_df["successes"] = model_df["n_whiffs"].astype(int)
model_df["trials"] = model_df["n_swings"].astype(int)
assert (model_df.successes <= model_df.trials).all()
assert model_df[PRED_VARS].notna().all().all()
```

## Model 1 — hierarchical linear beta-binomial

Swings are the trials and whiffs the successes. A plain binomial would assume
the only variation in whiff rate at fixed mechanics is sampling noise; the
histogram above shows far more spread than that, so the `beta_binomial` family
carries an extra dispersion parameter `kappa`.

The varying intercept **and** varying slopes by `pitch_name` are what make this
hierarchical: velocity should matter differently for a four-seamer than for a
changeup, and each pitch type gets a partially-pooled posterior of its own
rather than being forced onto one league-wide slope.

```{python}
#| label: py-m1-build
def hierarchical_formula(terms, group="pitch_name", response="p(successes, trials)"):
    fixed = " + ".join(terms)
    return f"{response} ~ {fixed} + (1 + {fixed} | {group})"

f1 = hierarchical_formula(PRED_VARS)
print(f1)

m1 = bmb.Model(f1, model_df, family="beta_binomial")
m1.build()
print(m1)

# model.graph() shells out to the graphviz binary, which is often not installed;
# it is a nicety, not a dependency.
try:
    m1.graph()
except Exception as exc:
    print(f"(skipping model.graph(): {type(exc).__name__}: {exc})")
```

```{python}
#| label: py-m1-fit
idata1 = m1.fit(**SAMPLE_KW)
ensure_log_likelihood(m1, idata1)

# NOTE: the response is literally named "p(successes, trials)", so the fitted
# mean does NOT land in idata.posterior["p_mean"]. bambi stores the
# beta-binomial's parameters under "mu" and "kappa".
RESPONSE_NAME = m1.response_component.term.name
print("response name:", RESPONSE_NAME)

az.summary(idata1, var_names=["Intercept", *PRED_VARS, "kappa"]).round(3)
```

# MCMC diagnostics

A posterior is only worth interpreting if the sampler actually explored it.
The checks below are the standard battery: split $\hat{R}$ (should be < 1.01),
bulk and tail effective sample size (should comfortably exceed 400 per
parameter), divergent transitions (should be zero), tree-depth saturation, and
E-BFMI (below ~0.3 indicates the sampler is struggling with the energy
distribution).

```{python}
#| label: py-diagnostics
def mcmc_report(idata, name):
    s = az.summary(idata)
    div = int(idata.sample_stats["diverging"].sum())
    n_draws = idata.posterior.sizes["chain"] * idata.posterior.sizes["draw"]
    bfmi = az.bfmi(idata)
    rows = {
        "model": name,
        "n_draws": n_draws,
        "max_rhat": round(float(s["r_hat"].max()), 4),
        "n_rhat_gt_1.01": int((s["r_hat"] > 1.01).sum()),
        "min_ess_bulk": int(s["ess_bulk"].min()),
        "min_ess_tail": int(s["ess_tail"].min()),
        "divergences": div,
        "div_pct": round(100 * div / n_draws, 3),
        "min_bfmi": round(float(np.min(bfmi)), 3),
    }
    if "tree_depth" in idata.sample_stats:
        md = int(idata.sample_stats.attrs.get("max_treedepth", 10))
        rows["pct_max_treedepth"] = round(
            100 * float((idata.sample_stats["tree_depth"] >= md).mean()), 2)
    worst = s.sort_values("ess_bulk").head(5)[["mean", "sd", "r_hat", "ess_bulk", "ess_tail"]]
    return pd.Series(rows), worst

rep1, worst1 = mcmc_report(idata1, "M1 linear")
print(rep1.to_string())
print("\nworst-mixing parameters:")
print(worst1.round(3).to_string())
```

```{python}
#| label: py-trace
az.plot_trace(idata1, var_names=["Intercept", *PRED_VARS[:3], "kappa"],
              compact=True, figsize=(11, 9))
plt.tight_layout()
plt.show()
```

```{python}
#| label: py-energy
az.plot_energy(idata1, figsize=(6, 4))
plt.show()
```

## Posterior predictive check

```{python}
#| label: py-ppc
# "pps" and "mean" are the deprecated spellings of these two; the current API
# is kind="response" (draws of the outcome) and kind="response_params" (mu, kappa).
m1.predict(idata1, kind="response", random_seed=RANDOM_SEED)
az.plot_ppc(idata1, num_pp_samples=200, figsize=(7, 4.5))
plt.title("PPC — whiff counts, M1")
plt.show()
```

```{python}
#| label: py-ppc-rate
# The PPC above is on the count scale, where differences in `trials` dominate.
# Re-express it as a whiff *rate* so it is comparable with the EDA histogram.
pp = idata1.posterior_predictive[RESPONSE_NAME]
trials = model_df["trials"].to_numpy()
pp_rate = (pp / trials).stack(sample=("chain", "draw")).transpose("sample", ...)

fig, ax = plt.subplots(figsize=(7, 4.5))
for i in rng.choice(pp_rate.sizes["sample"], 150, replace=False):
    ax.hist(pp_rate[i].values, bins=40, range=(0, 1), histtype="step",
            density=True, color="C0", alpha=0.06)
ax.hist(model_df["whiff_rate"], bins=40, range=(0, 1), histtype="step",
        density=True, color="k", lw=2, label="observed")
ax.set_xlabel("whiff rate"); ax.set_ylabel("density")
ax.set_title("PPC on the rate scale — M1"); ax.legend()
plt.show()
```

# Testing the linearity assumption

A beta-binomial GLM does not assume the whiff *rate* is linear in mechanics — it
assumes the **logit** of the rate is linear in them. That is the assumption to
test, and the right tool is the residual, not the raw scatter.

Two complementary checks:

**Binned residual plots.** Raw residuals from a binomial-type model are useless
to eyeball, so residuals are averaged within bins of each predictor (Gelman &
Hill). Under a correct specification the binned means scatter around zero with
no trend; roughly 95% should fall inside the $\pm 2\,\text{SE}$ band.

**A weighted RESET-style curvature test.** For each predictor $x$, regress the
residual on $x$ and $x^2$ (weighted by number of swings). A non-null quadratic
coefficient is direct evidence of curvature the linear term is not capturing.
A global version regresses the residual on the fitted linear predictor and its
square. Bonferroni-corrected across predictors.

```{python}
#| label: py-resid
import statsmodels.api as sm

m1.predict(idata1, kind="response_params")

def fitted_mu(idata):
    """Posterior-mean fitted probability per observation."""
    return idata.posterior["mu"].mean(dim=("chain", "draw")).values

mu1 = fitted_mu(idata1)
eta1 = logit(np.clip(mu1, 1e-6, 1 - 1e-6))
y_prop = (model_df["successes"] / model_df["trials"]).to_numpy()
resid1 = y_prop - mu1
w = model_df["trials"].to_numpy().astype(float)
print(f"mean residual: {np.average(resid1, weights=w):+.5f}  (should be ~0)")
```

```{python}
#| label: py-binned-resid
def binned_residuals(x, resid, weights, n_bins=40):
    """Gelman-Hill binned residuals: mean residual and +/-2SE band per x-quantile bin."""
    order = np.argsort(x)
    xs, rs, ws = x[order], resid[order], weights[order]
    edges = np.array_split(np.arange(len(xs)), n_bins)
    out = []
    for idx in edges:
        if len(idx) < 2:
            continue
        out.append((xs[idx].mean(), rs[idx].mean(),
                    2 * rs[idx].std(ddof=1) / np.sqrt(len(idx)), len(idx)))
    return pd.DataFrame(out, columns=["x", "resid", "se2", "n"])

fig, axes = plt.subplots(2, 3, figsize=(12, 7), sharey=True)
for ax, v in zip(axes.ravel(), PRED_VARS):
    b = binned_residuals(model_df[v].to_numpy(), resid1, w)
    ax.axhline(0, color="k", lw=0.8)
    ax.fill_between(b.x, -b.se2, b.se2, color="C0", alpha=0.15)
    ax.scatter(b.x, b.resid, s=14, color="C0")
    # LOESS through the binned means to make curvature visible
    sm_l = sm.nonparametric.lowess(b.resid, b.x, frac=0.6, return_sorted=True)
    ax.plot(sm_l[:, 0], sm_l[:, 1], color="crimson", lw=1.6)
    ax.set_title(v.replace("avg_", ""), fontsize=9)
    ax.set_xlabel("z-score")
for ax in axes.ravel()[len(PRED_VARS):]:
    ax.set_visible(False)
axes[0, 0].set_ylabel("mean residual"); axes[1, 0].set_ylabel("mean residual")
fig.suptitle("Binned residuals vs predictors — M1 (linear)")
plt.tight_layout()
plt.show()
```

```{python}
#| label: py-linearity-test
def curvature_tests(frame, preds, resid, weights):
    rows = []
    for v in preds:
        x = frame[v].to_numpy()
        X = sm.add_constant(np.column_stack([x, x ** 2]))
        fit = sm.WLS(resid, X, weights=weights).fit()
        rows.append({"term": v, "quad_coef": fit.params[2],
                     "std_err": fit.bse[2], "t": fit.tvalues[2],
                     "p_value": fit.pvalues[2]})
    out = pd.DataFrame(rows)
    out["p_bonferroni"] = np.minimum(1.0, out.p_value * len(out))
    out["nonlinear"] = out.p_bonferroni < 0.05
    return out.sort_values("p_value")

curv = curvature_tests(model_df, PRED_VARS, resid1, w)
print(curv.round(4).to_string(index=False))

# Global Ramsey RESET on the linear predictor.
X_reset = sm.add_constant(np.column_stack([eta1, eta1 ** 2, eta1 ** 3]))
reset_fit = sm.WLS(resid1, X_reset, weights=w).fit()
reset_p = float(reset_fit.f_test(np.eye(4)[[2, 3]]).pvalue)
print(f"\nglobal RESET (eta^2, eta^3) p-value: {reset_p:.4g}")

LINEARITY_OK = (not curv.nonlinear.any()) and (reset_p >= 0.05)
print(f"\nlinearity assumption satisfied: {LINEARITY_OK}")
if not LINEARITY_OK:
    print("violating terms:", curv.loc[curv.nonlinear, 'term'].tolist() or "(global only)")
```

## Follow-up EDA on the non-linear terms

Where the curvature test fires, look directly at the empirical logit — the
observed response on the model's own scale — against the predictor. The
half-continuity correction keeps groups with zero or all whiffs finite.

```{python}
#| label: py-emp-logit
# If the tests reject linearity, examine exactly the terms they rejected. If
# they do not, still carry the three most-curved terms forward: Model 2 is the
# nonlinear comparison candidate either way, and a spline model that fails to
# win is itself the evidence that the linear form is adequate.
if curv.nonlinear.any():
    FLAGGED = curv.loc[curv.nonlinear, "term"].tolist()
    print("linearity rejected for:", FLAGGED)
else:
    FLAGGED = curv.head(3)["term"].tolist()
    print("linearity not rejected; carrying most-curved terms as a check:", FLAGGED)

emp_logit = np.log((model_df.successes + 0.5) /
                   (model_df.trials - model_df.successes + 0.5)).to_numpy()

fig, axes = plt.subplots(1, len(FLAGGED), figsize=(4.2 * len(FLAGGED), 4), squeeze=False)
for ax, v in zip(axes.ravel(), FLAGGED):
    x = model_df[v].to_numpy()
    ax.scatter(x, emp_logit, s=6, alpha=0.15, color="grey")
    lo = sm.nonparametric.lowess(emp_logit, x, frac=0.4, return_sorted=True)
    ax.plot(lo[:, 0], lo[:, 1], color="crimson", lw=2, label="LOESS")
    b = np.polyfit(x, emp_logit, 1)
    xs = np.linspace(x.min(), x.max(), 50)
    ax.plot(xs, np.polyval(b, xs), color="navy", ls="--", lw=1.6, label="linear fit")
    ax.set_title(v.replace("avg_", "")); ax.set_xlabel("z-score")
    ax.legend(fontsize=8)
axes.ravel()[0].set_ylabel("empirical logit(whiff rate)")
fig.suptitle("Empirical logit vs flagged predictors")
plt.tight_layout()
plt.show()
```

```{python}
#| label: py-emp-logit-bytype
# The same curve, split by pitch type: a bend that appears only in the pooled
# view can be Simpson's paradox from differing pitch-type means rather than a
# genuine within-type non-linearity. If it survives the split, it is real.
v = FLAGGED[0]
types = model_df.pitch_name.value_counts().head(6).index.tolist()
fig, ax = plt.subplots(figsize=(7.5, 4.5))
for i, t in enumerate(types):
    sub = model_df[model_df.pitch_name == t]
    x = sub[v].to_numpy()
    el = np.log((sub.successes + 0.5) / (sub.trials - sub.successes + 0.5)).to_numpy()
    lo = sm.nonparametric.lowess(el, x, frac=0.7, return_sorted=True)
    ax.plot(lo[:, 0], lo[:, 1], lw=1.8, color=f"C{i}", label=t)
ax.set_xlabel(f"{v} (z-score)"); ax.set_ylabel("empirical logit(whiff rate)")
ax.set_title(f"Within-pitch-type shape of {v}"); ax.legend(fontsize=8)
plt.show()
```

# Model 2 — hierarchical spline beta-binomial

Where linearity fails, the fix is to keep the hierarchy and the beta-binomial
likelihood exactly as they are and relax only the *functional form*: replace the
flagged linear terms with a cubic B-spline basis. This is still fitted with
NUTS, so it remains a fully Bayesian MCMC model — it simply has a more flexible
mean function.

The group-specific terms stay linear. Per-pitch-type spline coefficients would
add `df` × (number of pitch types) parameters, and with a few dozen pitchers in
the smaller types there is not enough information to identify them; the model
would be all prior and no data.

```{python}
#| label: py-m2-build
SPLINE_DF = 4
spline_terms = [f"bs({v}, df={SPLINE_DF})" if v in FLAGGED else v for v in PRED_VARS]

f2 = (
    "p(successes, trials) ~ " + " + ".join(spline_terms)
    + " + (1 + " + " + ".join(PRED_VARS) + " | pitch_name)"
)
print(f2)

# B-spline basis columns are bounded in [0, 1] with small variance, so bambi's
# automatic prior scaling inflates their sigmas to 10-30 on the logit scale.
# That is effectively flat and lets the spline chase noise between knots. Effects
# here are order 1 on the logit scale, so give the spline coefficients an
# explicitly regularizing prior and pin the intercept and group-level scale to
# match M1 -- otherwise M2 wins or loses the comparison on prior width rather
# than on functional form.
priors2 = {t: bmb.Prior("Normal", mu=0, sigma=1.5)
           for t in spline_terms if t.startswith("bs(")}
priors2["Intercept"] = bmb.Prior("Normal", mu=0, sigma=2.5)
priors2["1|pitch_name"] = bmb.Prior(
    "Normal", mu=0, sigma=bmb.Prior("HalfNormal", sigma=2.5)
)

m2 = bmb.Model(f2, model_df, family="beta_binomial", priors=priors2)
m2.build()
print(m2)
```

```{python}
#| label: py-m2-fit
idata2 = m2.fit(**SAMPLE_KW)
ensure_log_likelihood(m2, idata2)
rep2, worst2 = mcmc_report(idata2, "M2 spline")
print(pd.concat([rep1, rep2], axis=1).to_string())
print("\nworst-mixing parameters (M2):")
print(worst2.round(3).to_string())
```

```{python}
#| label: py-m2-resid
m2.predict(idata2, kind="response_params")
mu2 = fitted_mu(idata2)
resid2 = y_prop - mu2

curv2 = curvature_tests(model_df, PRED_VARS, resid2, w)
print("curvature tests after splines:")
print(curv2.round(4).to_string(index=False))

fig, axes = plt.subplots(1, len(FLAGGED), figsize=(4.2 * len(FLAGGED), 3.6), squeeze=False)
for ax, v in zip(axes.ravel(), FLAGGED):
    b = binned_residuals(model_df[v].to_numpy(), resid2, w)
    ax.axhline(0, color="k", lw=0.8)
    ax.fill_between(b.x, -b.se2, b.se2, color="C2", alpha=0.15)
    ax.scatter(b.x, b.resid, s=14, color="C2")
    lo = sm.nonparametric.lowess(b.resid, b.x, frac=0.6, return_sorted=True)
    ax.plot(lo[:, 0], lo[:, 1], color="crimson", lw=1.6)
    ax.set_title(v.replace("avg_", ""), fontsize=9)
fig.suptitle("Binned residuals — M2 (splines on flagged terms)")
plt.tight_layout()
plt.show()
```

## Fit comparison by LOO

```{python}
#| label: py-loo
comparison = az.compare({"M1 linear": idata1, "M2 spline": idata2}, ic="loo")
print(comparison.round(2).to_string())
az.plot_compare(comparison, figsize=(7, 3))
plt.show()
```

LOO measures out-of-sample density for the *whiff* response. It is a useful fit
check, but it is not the criterion asked for — that is run value, below.

# Per-pitch-type posteriors

The hierarchy gives every pitch type its own posterior. Two views of it: the
posterior for each type's *slope* on a mechanic (does velocity buy more whiffs
on a four-seamer than on a changeup?), and the posterior for each type's whiff
rate at that type's own average mechanics.

```{python}
#| label: py-forest-slopes
FOCUS = "avg_release_speed"
az.plot_forest(
    idata1,
    var_names=[FOCUS, f"{FOCUS}|pitch_name"],
    combined=True, figsize=(7.5, 5.5), r_hat=True,
)
plt.title(f"Posterior slope on {FOCUS}: population and by pitch type")
plt.tight_layout()
plt.show()
```

```{python}
#| label: py-type-posterior
def type_posterior(model, idata, frame, preds, group="pitch_name"):
    """Posterior whiff rate for each pitch type, evaluated at that type's own
    average mechanics (all predictors are standardized, so this is the type's
    typical pitch rather than a league-average one)."""
    grid = (frame.groupby(group, observed=True)[preds].mean().reset_index())
    grid["successes"], grid["trials"] = 0, 1
    pred = model.predict(idata, data=grid, kind="response_params", inplace=False)
    mu = pred.posterior["mu"]
    mu = mu.assign_coords({mu.dims[-1]: grid[group].to_numpy()})
    return grid, mu

grid1, mu_by_type1 = type_posterior(m1, idata1, model_df, PRED_VARS)

summary_by_type = (
    az.summary(mu_by_type1, hdi_prob=0.94)
    .assign(pitch_name=grid1["pitch_name"].to_numpy())
    .set_index("pitch_name")[["mean", "sd", "hdi_3%", "hdi_97%"]]
    .sort_values("mean", ascending=False)
)
print(summary_by_type.round(3).to_string())

az.plot_forest(mu_by_type1, combined=True, figsize=(7, 4.5))
plt.title("Posterior whiff rate at pitch-type-average mechanics (M1)")
plt.tight_layout()
plt.show()
```

# Stuff+

**Definition.** For posterior draw $d$, pitcher $i$ throwing pitch type $k$, let
$\eta_{id}$ be the model's latent log-odds of a whiff. Standardize it against
the posterior for *that pitch type*:

$$
z_{id} \;=\; \frac{\eta_{id} - m_{kd}}{s_{kd}},
\qquad
\text{Stuff+}_{id} \;=\; 100 + 10\,z_{id}
$$

where $m_{kd}$ and $s_{kd}$ are the mean and standard deviation of $\eta_{\cdot d}$
across the pitchers who throw type $k$, computed **within each draw**. So 100 is
average for that pitch type, one standard deviation is 10 points, and higher is
better. Doing the standardization draw-by-draw rather than on posterior means
means Stuff+ inherits the model's uncertainty: each pitcher gets a posterior
*distribution* of Stuff+, not a point estimate.

```{python}
#| label: py-stuffplus
MAX_DRAWS = 2000

def stuff_plus_draws(idata, frame, group="pitch_name", center=100.0, scale=10.0,
                     max_draws=MAX_DRAWS):
    mu = (idata.posterior["mu"]
          .stack(sample=("chain", "draw"))
          .transpose("sample", ...)
          .values)
    if mu.shape[0] > max_draws:                       # thin for memory
        idx = np.linspace(0, mu.shape[0] - 1, max_draws).astype(int)
        mu = mu[idx]
    eta = logit(np.clip(mu, 1e-9, 1 - 1e-9))
    codes = pd.Categorical(frame[group]).codes
    z = np.empty_like(eta)
    for k in np.unique(codes):
        sel = codes == k
        sub = eta[:, sel]
        z[:, sel] = (sub - sub.mean(axis=1, keepdims=True)) / sub.std(axis=1, ddof=1, keepdims=True)
    return center + scale * z            # (draws, observations)

def summarise_stuff(sp_draws, frame):
    # az.hdi expects (chain, draw, *shape); sp_draws is (draw, obs).
    hdi = az.hdi(sp_draws[None, :, :], hdi_prob=0.94)  # -> (obs, 2)
    hdi = np.asarray(hdi["x"] if hasattr(hdi, "keys") else hdi)
    return frame.assign(
        stuff_plus=sp_draws.mean(axis=0),
        stuff_plus_sd=sp_draws.std(axis=0, ddof=1),
        stuff_plus_lo=hdi[:, 0],
        stuff_plus_hi=hdi[:, 1],
    )

sp1 = stuff_plus_draws(idata1, model_df)
sp2 = stuff_plus_draws(idata2, model_df)

stuff1 = summarise_stuff(sp1, df.copy())
stuff2 = summarise_stuff(sp2, df.copy())

print(stuff1.groupby("pitch_name", observed=True)["stuff_plus"]
      .agg(["mean", "std", "min", "max"]).round(2).to_string())
```

```{python}
#| label: py-stuff-leaders
cols = ["player_name", "pitch_name", "n_pitches", "whiff_rate",
        "stuff_plus", "stuff_plus_sd", "stuff_plus_lo", "stuff_plus_hi", "run_value"]
print("top 15 by Stuff+ (M1)")
print(stuff1.sort_values("stuff_plus", ascending=False).head(15)[cols]
      .round(3).to_string(index=False))
print("\nbottom 10 by Stuff+ (M1)")
print(stuff1.sort_values("stuff_plus").head(10)[cols].round(3).to_string(index=False))
```

```{python}
#| label: py-stuff-uncertainty
fig, axes = plt.subplots(1, 2, figsize=(11, 4))
axes[0].hist(stuff1.stuff_plus, bins=40, color="steelblue", edgecolor="white")
axes[0].axvline(100, color="red", ls="--")
axes[0].set_xlabel("Stuff+"); axes[0].set_ylabel("count")
axes[0].set_title("Stuff+ distribution (M1)")

axes[1].scatter(stuff1.n_pitches, stuff1.stuff_plus_sd, s=8, alpha=0.35)
axes[1].set_xscale("log")
axes[1].set_xlabel("pitches thrown (log)"); axes[1].set_ylabel("posterior SD of Stuff+")
axes[1].set_title("Uncertainty shrinks with sample size")
plt.tight_layout()
plt.show()
```

# Model selection on cumulative run value

The criterion is **$R^2$ and MSE for cumulative run value**. Two things have to
be handled for that comparison to mean anything.

*Exposure.* `run_value` is a **sum** over pitches, so it scales with usage: a
pitcher who threw 600 sliders has a larger $|{\rm run\ value}|$ than one who
threw 40, regardless of quality. Any evaluation that ignores `n_pitches` mostly
measures playing time. Pitch count is therefore included in every specification,
including the baseline, so the reported gain is Stuff+'s **incremental**
contribution rather than a restatement of usage.

*Leakage.* Rows are pitcher × pitch-type, and one pitcher contributes several
rows. Splitting rows at random would put the same pitcher's talent on both sides
of the fold. Cross-validation is grouped by pitcher.

Scores are reported out-of-sample; in-sample values are shown alongside only to
expose overfitting.

```{python}
#| label: py-eval
from sklearn.linear_model import LinearRegression
from sklearn.metrics import mean_squared_error, r2_score
from sklearn.model_selection import GroupKFold

TARGET = "run_value"          # cumulative run value, negative favours pitcher
N_SPLITS = 5
BASELINE = "baseline (exposure + pitch type)"

def design(frame, use_stuff):
    """Exposure + pitch-type fixed effects, optionally plus Stuff+."""
    X = pd.get_dummies(frame[["pitch_name"]], drop_first=True, dtype=float)
    X["n_pitches"] = frame["n_pitches"].to_numpy()
    if use_stuff:
        X["stuff_plus"] = frame["stuff_plus"].to_numpy()
        # Stuff+ is standardized within pitch type, so its effect on a *cumulative*
        # total must scale with how often the pitch was thrown.
        X["stuff_x_pitches"] = frame["stuff_plus"].to_numpy() * frame["n_pitches"].to_numpy()
    return X.to_numpy()

def evaluate(frame, label, use_stuff=True, target=TARGET):
    X, y = design(frame, use_stuff), frame[target].to_numpy()
    groups = frame["player_name"].to_numpy()
    oof = np.zeros_like(y, dtype=float)
    for tr, te in GroupKFold(n_splits=N_SPLITS).split(X, y, groups):
        oof[te] = LinearRegression().fit(X[tr], y[tr]).predict(X[te])
    ins = LinearRegression().fit(X, y).predict(X)
    return {
        "model": label,
        "cv_r2": r2_score(y, oof),
        "cv_mse": mean_squared_error(y, oof),
        "cv_rmse": np.sqrt(mean_squared_error(y, oof)),
        "in_sample_r2": r2_score(y, ins),
        "in_sample_mse": mean_squared_error(y, ins),
    }

results = pd.DataFrame([
    evaluate(stuff1, BASELINE, use_stuff=False),
    evaluate(stuff1, "M1 linear -> Stuff+"),
    evaluate(stuff2, "M2 spline -> Stuff+"),
]).set_index("model")

print("target: cumulative run value")
print(results.round(4).to_string())
```

```{python}
#| label: py-eval-rate
# Secondary view on the rate scale, where exposure is already divided out.
results_rate = pd.DataFrame([
    evaluate(stuff1, "baseline (pitch type)", use_stuff=False, target="rv_per_100"),
    evaluate(stuff1, "M1 linear -> Stuff+", target="rv_per_100"),
    evaluate(stuff2, "M2 spline -> Stuff+", target="rv_per_100"),
]).set_index("model")

print("target: run value per 100 pitches")
print(results_rate.round(4).to_string())
```

```{python}
#| label: py-select
# For a fixed target, R^2 = 1 - MSE / Var(y). The two criteria are monotone
# transforms of one another and therefore always produce the same ranking --
# reporting both is still worthwhile because MSE is on the scale of runs^2 and
# says how large the errors actually are, but it cannot break a tie. The real
# question is whether the leader's margin justifies the extra parameters.
candidates = results.drop(index=BASELINE)
ranked = candidates.sort_values("cv_r2", ascending=False)
print(ranked[["cv_r2", "cv_mse", "cv_rmse"]].round(4).to_string())
assert ranked.index[0] == candidates.cv_mse.idxmin(), "R^2 and MSE must agree"

PARSIMONY_TOL = 0.005
leader = ranked.index[0]
margin = float(ranked.cv_r2.iloc[0] - ranked.cv_r2.iloc[1])

if leader.startswith("M2") and margin < PARSIMONY_TOL:
    SELECTED = ranked.index[1]
    print(f"\n{leader} leads by only {margin:.4f} R^2, inside the "
          f"{PARSIMONY_TOL} tolerance -- keeping the simpler linear model.")
else:
    SELECTED = leader
    print(f"\n{leader} leads by {margin:.4f} R^2.")

print(f"SELECTED: {SELECTED}")
print(f"LOO preference (whiff fit, for reference): {comparison.index[0]}")

best_stuff = stuff1 if SELECTED.startswith("M1") else stuff2
gain = candidates.loc[SELECTED, "cv_r2"] - results.loc[BASELINE, "cv_r2"]
print(f"incremental out-of-sample R^2 from Stuff+: {gain:+.4f}  "
      f"(baseline {results.loc[BASELINE, 'cv_r2']:.4f} -> "
      f"{candidates.loc[SELECTED, 'cv_r2']:.4f})")
```

```{python}
#| label: py-eval-plot
fig, ax = plt.subplots(figsize=(6.5, 5))
sc = ax.scatter(best_stuff.stuff_plus, best_stuff.rv_per_100,
                c=best_stuff.n_pitches, s=14, alpha=0.6, cmap="viridis",
                norm=plt.matplotlib.colors.LogNorm())
b = np.polyfit(best_stuff.stuff_plus, best_stuff.rv_per_100, 1)
xs = np.linspace(best_stuff.stuff_plus.min(), best_stuff.stuff_plus.max(), 50)
ax.plot(xs, np.polyval(b, xs), color="crimson", lw=2)
ax.axhline(0, color="k", lw=0.8); ax.axvline(100, color="k", lw=0.8, ls=":")
plt.colorbar(sc, label="pitches thrown")
ax.set_xlabel("Stuff+"); ax.set_ylabel("run value per 100 pitches")
ax.set_title(f"Stuff+ vs run value — {SELECTED}")
r = np.corrcoef(best_stuff.stuff_plus, best_stuff.rv_per_100)[0, 1]
ax.annotate(f"r = {r:.3f}", xy=(0.04, 0.06), xycoords="axes fraction")
plt.tight_layout()
plt.show()
```

```{python}
#| label: py-export
out_cols = [
    "player_name", "pitch_name", "n_pitches", "n_swings", "n_whiffs",
    "whiff_rate", "run_value", "rv_per_100",
    "stuff_plus", "stuff_plus_sd", "stuff_plus_lo", "stuff_plus_hi",
]
best_stuff.sort_values(["pitch_name", "stuff_plus"], ascending=[True, False]) \
    [out_cols].to_csv("data/stuff_plus.csv", index=False)

pd.concat({"cumulative_run_value": results, "run_value_per_100": results_rate}) \
    .to_csv("data/model_selection.csv")

print("wrote data/stuff_plus.csv and data/model_selection.csv")
print(f"selected model: {SELECTED}")
```

# Summary

| Stage | Result |
|---|---|
| Missing data | `arm_angle` is absent for some pitchers entirely; incomplete rows dropped explicitly before standardizing |
| Collinearity | VIF pruning at threshold 5 — `arm_angle` is near-deterministic in release position |
| Overdispersion | Beta-binomial over binomial, `kappa` estimated |
| Hierarchy | Varying intercept + slopes by `pitch_name`; each type gets its own posterior |
| Linearity | Tested by binned residuals + weighted curvature tests; splines fitted where it fails |
| MCMC | $\hat{R}$, ESS bulk/tail, divergences, tree depth, E-BFMI, energy plot, PPC |
| Stuff+ | Draw-wise z-score against the pitch type's own posterior, 100 ± 10 |
| Selection | Grouped 5-fold CV $R^2$ / MSE on cumulative run value, with exposure in the baseline |
